---
title: "ML-Final project"
author: "JOMR"
date: "8/7/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Using Machine Learning to classify activities using mobile devices  
The purpose of this project is to use machine learning to categorize user´s activities.  

```{r reading, cache=TRUE, warning=FALSE}
library(tidyverse)
library(caret)
library(kableExtra)
#setwd("final project")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile="training.csv")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile="test.csv")
train <-read.csv("training.csv", header=TRUE)
test <- read.csv("test.csv", header=TRUE)
```   
We will first clean up the database; so that only the variables that are present (ie. have no missings) in both databases are used in the prediction and the classification algorhitms.  
```{r cleaning, cache=TRUE}
#De-selecting columns that have missing values
whichcleantraining <- which(apply(train, 2, function(x) sum(is.na(x))) == 0 )
traincleaned <- train[,whichcleantraining]
whichcleantesting <- which(apply(test, 2, function(x) sum(is.na(x))) == 0 )
testcleaned <- test[,whichcleantesting]
#intersecting names on each db
names1 <- colnames(testcleaned)
names2 <- colnames(traincleaned)
both <- names1[names1 %in% names2]
#databases with enough variables
traindb <- dplyr::select(train,c(both,"classe"))
testdb <- dplyr::select(test,c(both,"problem_id"))
#finding variables with low variance and unusable variables and eliminating them
nearZeroVar(traindb) #variable is "new_window"
traindb <- select(traindb, -c(new_window, X, user_name, cvtd_timestamp, raw_timestamp_part_1, raw_timestamp_part_2, num_window))
testdb <- select(testdb, -c(new_window, X, user_name, cvtd_timestamp, raw_timestamp_part_1, raw_timestamp_part_2, num_window))
```   
## Machine learning  
I will proceed to build the classification algorithm in the following way:   
* Build an rpart tree  
* Build a random forest model  
* Cross-validate both approaches using k-folds cross-validation  
* Predict using the best of the two models in the test dataset.   
```{r ml1, cache=TRUE}
library(caret)
#setting cross-validation parameters
train_control <- trainControl(method="cv", number=10, classProbs = T)
#Using rpart method to build a classification tree
rpart1 <- train(classe ~ ., data=traindb, method="rpart", trControl=train_control)
#using a random forest model
rforest1 <- train(classe ~., data=traindb, method="rf",trControl=train_control)
#using k folds cv
print(rpart1) # expected accuracy 50% (50% out-of-sample error)
print (rforest1) #expected accuracy close to 99% (1% out-of sample error)
```  
Since the out-of-sample error for rforest model has a higher accuracy, I choose the random forest-based prediction for the final submission.  
The following are the predictions associated with the model and the test db:
`r predict(rforest1, newdata=testdb)` .  